---
title: 'Unsupervised Learning: PSet 3'
author: "Felipe Alamos"
date: "10/24/2019"
output: pdf_document
---

```{r, message=FALSE}
#Setups

library(dplyr)
#library(knitr)
library(skimr)
library(seriation)
library(ggplot2)
library(dbscan)
library(mixtools)
library(plotGMM)
library(gridExtra)
```

#1.
_Load the state legislative professionalism data from the relevant subdirectory in this repo. See the codebook for reference in the same subdirectory and combine with our discussion of these data and the concept of state legislative professionalism from class for relevant background information._

```{r}
data <- get(load("StateLegProfData&Codebook/legprof-components.v1.0.RData"))
head(data)
```

#2.
_Munge the data:_
* select only the continuous features that should capture a state legislature's level of "professionalism" (session length (total and regular), salary, and expenditures); 
* restrict the data to only include the 2009/10 legislative session for consistency; 
* omit all missing values; 
* standardize the input features;
* and anything else you think necessary to get this subset of data into workable form (_hint_: consider storing the state names as a separate object to be used in plotting later)

```{r}
data_filtered <- data %>%
  filter(sessid == "2009/10") %>%
  na.omit() %>%
  select(t_slength, slength, salary_real, expend)

#Standardizing
data_scaled <- as.data.frame(scale(data_filtered))

states <- data %>%
  filter(sessid == "2009/10") %>%
  na.omit() %>%
  select(stateabv)
```


#3.
_Perform quick EDA visually or numerically and discuss the patterns you see._

```{r}
#skim(data_filtered), some problem with Knit happening
summary(data_filtered)
```

- Most of the data is from 2010
- All numerical variables are skewed to the left, i.e, they have significant outliers on their high end. This is particularly exacerbated for expenditure, where its max value is more than 100 times the median value. Lets have a closer look at expend:

```{r}
data_filtered_with_state<-data_filtered
data_filtered_with_state$state <- states$stateabv

ggplot(data_filtered_with_state, aes(x=reorder(as.factor(state),-expend), expend)) +
  geom_bar(stat="identity",color="black", fill="white")+
   theme(axis.text.x = element_text(size=9, angle = 90, vjust=-0.001))+
   labs(x = "States")
```

- We observe that the state with the extreme `expenditure` value is CA.
- We know try to understand correlation between the variables

#session length (total and regular), salary, and expenditures
```{r}
#stateabv, year, t_slength, slength, salary_real, expend
ggplot(data_scaled, aes(x=slength, y=salary_real)) +
  geom_point()
ggplot(data_scaled, aes(x=slength, y=expend)) +
  geom_point()
```

- We observe that there seems to be a positive correlation between length of regular session and salary. We might state the same for expenditure, but the correlation is much weaker.

# 4

_Diagnose clusterability in any way you'd prefer (e.g., sparse sampling, ODI, etc.); display the results and discuss the likelihood that natural, non-random structure exist in these data._

```{r}

df_dist<- dist(data_scaled)
dissplot(df_dist)
```
- From a quick look at this ODI, it seems that there is one big cluster, and then a second small one
- It does look that some outliers exist - elements that are very far away in distance from most others (this might be the ones groupin in the small cluster).

# 5
_Fit a k-means algorithm to these data and present the results. Give a quick, high level summary of the output and general patterns. Initialize the algorithm at k=2, and then check this assumption in the validation questions below._

```{r}
set.seed(123)
kmeans <- kmeans(data_scaled, 
              centers = 2,
              nstart = 15)

str(kmeans)

kmeans$size
```

- We observe two clusters, one with 43 elements and the other with 6.
- We present the centers of the two clusters:
```{r}
kmeans$centers
```
-And the list indicating to which cluster was each state associated
```{r}
kmeans$cluster

#We create a df where to save the clusters
data_scaled_with_clusters <- data_scaled
data_scaled_with_clusters$KmeansCluster <- as.factor(kmeans$cluster)
```

# 6.

_Fit a Gaussian mixture model via the EM algorithm to these data and present the results. Give a quick, high level summary of the output and general patterns. Initialize the algorithm at k=2, and then check this assumption in the validation questions below._

```{r}


#We use the multi-variate method so as to consider all columns of data
set.seed(123)
gmm1 <- mvnormalmixEM(data_scaled, k = 2) 
```

For the two cluster, we observe means on each of the variables:
```{r}
gmm1$mu
```
...standard deviations (notice that here we present the covariance matrix - in the diagonal we can find standard deviations)
```{r}
gmm1$sigma
```
... and heights:
```{r}
gmm1$lambda
```

... and the probabilities of each point being part of each of the two clusters:
```{r}
post <- as.data.frame(gmm1$posterior)
post

# quick viz distribution of mixture proportions
plot(post$comp.1, post$comp.2) 
```
- We can observe that states are represented either by one distribution or the other, not a mixture of them

```{r}
# get counts for each component 
post$component <- ifelse(post$comp.1 < 0.2, 2, 1)
table(post$component) 
```
- The first cluster has 44 states, the sencond one 5.

```{r}
# We add the clusters obtained by gmm to our dataframe
data_scaled_with_clusters$ClusterGMM <- post$component
```

# 7.
Fit one additional partitioning technique of your choice (e.g., PAM, CLARA, fuzzy C-means, DBSCAN, etc.), and present and discuss results. Here again initialize at k=2.

We will be using DBSCAN:
```{r}


#To identify the ideal epsilon, we can follow the techinque explained here: http://www.sthda.com/english/wiki/wiki.php?id_contents=7940 Nonetheless, we will use an epsilon that can generate 2 clusters, so as to compare it with the other algorithms.
#dbscan::kNNdistplot(data_scaled, k =  5)
#abline(h = 0.15, lty = 2)
#knee ideal ~1.5

dbscan <- dbscan(data_scaled, 
                    eps= 0.5) 
dbscan$cluster #0 indicates noise point

n_non_clustered <- sum(dbscan$cluster == 0)
n_clustered_in_1<- sum(dbscan$cluster == 1)
n_clustered_in_2<- sum(dbscan$cluster == 2)
```

-We observe that the 49 datapoints have been separated in two clusters. One with `r n_clustered_in_1` states, the other with `r n_clustered_in_2` states, and `r n_non_clustered` were not clustered.

```{r}
data_scaled_with_clusters$ClusterDBscan <- dbscan$cluster
```

# 8.

_Compare output of all in a visually useful, simple way (e.g., plotting by state cluster assignment across two features like salary and expenditures)._

```{r}
data_scaled_with_clusters

# View the salary by clusters
kmeans_salary_plot <- ggplot(data_scaled_with_clusters, aes(x = salary_real)) + 
  geom_histogram(aes(fill = factor(KmeansCluster)), stat ="bin", binwidth = 3) +
  labs(x = "Standarized salary",
       y = "Count of States",
       title = "KMeans Mixture Model") +
  scale_fill_manual(values=c("chartreuse4", "dodgerblue4"),
                    name="Component") +
  theme_bw()

gmm_salary_plot <- ggplot(data_scaled_with_clusters, aes(x = salary_real)) + 
  geom_histogram(aes(fill = factor(ClusterGMM)), stat ="bin", binwidth = 3) +
  labs(x = "Standarized salary",
       y = "Count of States",
       title = "Gaussian Mixture Model") +
  scale_fill_manual(values=c("chartreuse3", "dodgerblue3"),
                    name="Component") +
  theme_bw()

dbscan_salary_plot <- ggplot(data_scaled_with_clusters, aes(x = salary_real)) + 
  geom_histogram(aes(fill = factor(ClusterDBscan)), stat ="bin", binwidth = 3) +
  labs(x = "Standarized salary",
       y = "Count of States",
       title = "DBScan Model") +
  scale_fill_manual(values=c("gainsboro", "chartreuse2", "dodgerblue1"),
                    name="Component") +
  theme_bw()

# view side by side
grid.arrange(kmeans_salary_plot, gmm_salary_plot, dbscan_salary_plot, ncol = 3)
```





9. Select a single validation strategy (e.g., compactness via min(WSS), average silhouette width, etc.), and calculate for all three algorithms. Display and compare your results for all three algorithms you fit (k-means, GMM, X). 

10. Discuss the validation output. 
* What can you take away from the fit? 
* Which approach is optimal? And optimal at what value of k? 
* What are reasons you could imagine selecting a technically "sub-optimal" partitioning method, regardless of the validation statistics? 


**dbscan:
The benefit of this approach is its flexible to find interesting, unique
configurations that could more accurately characterize complex space
Regardless, the ability to not cluster each observation makes this
approach relatively free from the influence of outliers (which could be
good or bad, depending on the data, domain, etc.)
Unlike to K-means, DBSCAN does not require the user to specify the number of clusters to be generated
DBSCAN can find any shape of clusters. The cluster doesn’t have to be circular.
DBSCAN can identify outliers